Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-06 exp1/iswc2013.tex
small changes spellcheck
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@518 43ef1190-90fd-4dd7-a769-d0121f02b80e

However, there are many of problems with this approach when used to
evaluate assertions about the overall performance of a system or
several systems: 
\begin{itemize}
\item As the complexity increases, the number of parameters that
one may wish to test increases dramatically.
\item Nothing compels anyone to find microbenchmarks that
  can refute an assertion, and there is no structured approach to
  investigate whether a benchmark is flawed.
\item There is no meaningful summary of the performance as a whole, so
  that an assertion about performance cannot be refuted as an
  optimization may have dentrimental side effects that cannot be
  identified, or cannot be compared to possibly positive effects in a
  different microbenchmark.
\end{itemize}

@@

Thus, attacking the complexity problem by eliminating parameters we
think influence the outcome for simplicity is indefensible as
scientific practice. We must instead tackle this problem by enhancing
our capability of testing complex systems and to identify situations
where our assumptions are flawed. Such constraints can only be made
when we have very good reason to suspect that a certain parameter does
not have any influence.

In this paper, we employ techniques from the well-established field in
statistics of Design of Experiments to point out a direction that
promises to address these problems. First, we design a simple
8-parameter experiment, and then we demonstrate how a full
factorial experiment can meaningfully provide a summary of the
statistics.

Then, we demonstrate how a fractional factorial experiment can be used
to maintain a structured approach to the problem while reducing the
number of runs the experiment needs, and we discuss the tradeoffs
involved. We then continue to discuss techniques that shows our
simplistic experiment to be flawed, yet scientifically better than the
practice of benchmarking. Finally, we outline a roadmap to establish
the use of these techniques in SPARQL endpoint evaluations both as a
practical tool and as a scientific more rigorous methodology.

@@

As the focus of this paper is didactical, the actual measurements are
much simpler than what has been used in benchmarking. We have used the
dataset of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
smallest of their datasets (that we found to be more than 15
MTriples). Moreover, we have taken subsets of that dataset by using
the last 1 or 2 MTriples from their file.

@@

We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
dataset, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
where one machine runs GNU/Linux Debian Squeeze, has 16 GB RAM, an
Intel Core2 Duo E8500 CPU and two Parallell-ATA disks in a RAID-1
configuration. The other machine runs Debian Wheezy, has 8 GB RAM, an
Intel Core i7-3520M CPU and a single SSD on SATA-III.

@@

To proceed beyond a visual judgement of significance, we may use
Lenth's method\todo{ref Lenth}, which is a simple method for
estimating significance in an experiment with no replicates. We plot
the results in figure~\ref{fig:fulllenth}. In
figure~\ref{fig:fullnormal}, Lenth's method is used to label the
significant effects. Visually, this is not as useful, since the large
number\todo{really include?} We see that we have very significant
effects. We also note that ``Implement'' itself is significantly
negative. We also note that the ``Implement:Lang'' interaction is
significantly positive, as expected from our setup, where ``Lang'' was
the detrimental side-effect we simulated.

To investigate whether the detrimental side-effects cancels the
positive effect of our simulated optimization, we have to formulate a
hypothesis test. By inspecting the normal plot in
figure~\ref{fig:fullnormal}, we see that only five factors are highly
significant by either being a significant main effect or participate
in a highly significant interaction. We say that the three factors
``BGPComp'', ``Range'' and ``Machine'' are \emph{inactive} since they
do not contribute significantly to the observed variation in
performance. In practical situations, the four factors ``Lang'',
``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
i.e. we cannot change the size of the dataset without loss of
information, or change the query, since it gives a different
answer. In practice, we can only control ``Implement'', i.e. we can
only change the implementation. In this context, we call ``Implement''
a \emph{control} factor and the other four active factors
\emph{environmental}.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are critical.\todo{add
  table of effects} That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. This should cause some alarm. It is a fundamental
property of two-level experiments that it can only model linear
effects. If the effect is truly linear, and the variance is small, the
two levels can be quite close together and still yield a reasonable
estimate for the effect. However, if the effect is non-linear, the
choice of levels becomes critical. In this case, the experimenter
should be concerned that they may have a non-linear effect that is not
well represented by the sizes of datasets. Such reasoning should be
applied to all significant effects.

@@

As mentioned, the full factorial experiment goes as $2^n$ and becomes
prohibitivly expensive when $n$ is a large number of
factor. As the evaluation of SPARQL endpoints is an inherently complex
problem, a large number of factors are needed, and so full experiments
cannot scale. However, we may reduce the size of the experiment
significantly by sacrificing some explanatory power, in the form of
fractional factorial experiments.

@@

For any other purpose than a rough idea of some key influencers on the
overall performance of the endpoint for the given levels, this little
experiment is insufficient.

We turn to the 64-run experiment and its normal plot in
figure~\ref{fig:frac64normalalpha005}. We see that the we get more significant
effects, and that they correspond well to those we saw in the full
factorial experiment. They also all have an intuitive
explanation. However, we do not see the effects that we found
worrisome in the full factorial design. The ``Implement:TripleC''
interaction emerges only for a level $\alpha=0.15$, see
figure~\ref{fig:frac64normalalpha015}. To ensure that we discover such
cases will be a key challenge in further studies.

@@

The importance of finding flaws in the levels is highlighted by the
normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
experiment made it feasible to run with a larger dataset, this
experiment uses the original 15~MTriples dataset of
\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
previous discussion, for larger ``TripleC'', our simulated performance
is actually worse, as evident from the positive
``Implement''. We also see that the interactions ``Implement:Lang''
and ``Implement:TripleC'' is highly significant. It is clear what
happened: For larger ``TripleC'', the delay we introduced in the
langMatches function has a devastating effect, since it delays the
execution for every returned triple, and thereby completely dominates
the execution time.

Moreover, we note that the ``Machine'' factor has become highly
significant in the other end along with its ``Machine:TripleC''
interaction. One should take careful notice that the ``Machine''
factor encompasses many different things, that should be different
factors; CPU, disks, available RAM, installed software, tuning
parameters and so on. It is advantegous to have such broad factors in
the design in early experiments, because they may point out unexpected
effects that should be further scrutinized. The ``Machine'' factor is
a good example of usage that also helps us manage
complexity: If possible, one may group many factors into one if one is
in doubt of its relevance and break it up in subsequent
experiments. In this case, the cause of the performance problem was
highly audible: It was the old P-ATA disks of the larger machine, but
in general, such clues are not available to the experimenter.

We are forced to conclude that the three preceeding experiments are
severely flawed, as they failed to take into account the non-linear
effects arising from larger data sizes on slow hardware.

Note that we do not take advantage of the usual performance of 4store
in this simulation, to the contrary, we chose 4store because of its
relative simplicity to modify it to suite our needs for illustrative
purposes, which consists of degrading it in many ways. The small
datasets used here does not imply that the approach will not work for
large datasets; properly configured, actual optimizations will give
very different runtimes, and the small number of runs will make it
possible to evaluate a very large number of factors.

@@

\section*{Acknowledgements}

>>>>

However, there are many of problems with this approach when used to
evaluate assertions about the overall performance of a system or
several systems: 
\begin{itemize}
\item As the complexity increases, the number of parameters that
one may wish to test increases dramatically.
\item Nothing compels anyone to find microbenchmarks that
  can refute an assertion, and there is no structured approach to
  investigate whether a benchmark is flawed.
\item There is no meaningful summary of the performance as a whole, so
  that an assertion about performance cannot be refuted as an
  optimization may have detrimental side effects that cannot be
  identified, or cannot be compared to possibly positive effects in a
  different microbenchmark.
\end{itemize}

@@

Thus, attacking the complexity problem by eliminating parameters we
think influence the outcome for simplicity is untenable as
scientific practice. We must instead tackle this problem by enhancing
our capability of testing complex systems and to identify situations
where our assumptions are flawed. Such constraints can only be made
when we have very good reason to suspect that a certain parameter does
not have any influence.

In this paper, we employ techniques from the well-established field in
statistics known as Design of Experiments (DoE) to point out a
direction that promises to address these problems. First, we design a
simple 8-parameter experiment, and then we demonstrate how a full
factorial experiment can meaningfully provide a summary of the
statistics.

Then, we demonstrate how a fractional factorial experiment can be used
to maintain a structured approach to the problem while reducing the
number of runs the experiment needs, and we discuss the trade-offs
involved. We then continue to discuss techniques that shows our
simplistic experiment to be flawed, yet scientifically better than the
practice of benchmarking. Finally, we outline a road-map to establish
the use of these techniques in SPARQL endpoint evaluations both as a
practical tool and as a scientifically more rigorous methodology.

@@

As the focus of this paper is didactical, the actual measurements are
much simpler than what has been used in benchmarking. We have used the
data-set of DBPedia SPARQL Benchmark\cite{mxro:Morsey2011DBpedia}, but only consider the
smallest of their data-sets (that we found to be more than 15
MTriples). Moreover, we have taken subsets of that data-set by using
the last 1 or 2 MTriples from their file.

@@

We have chosen 8 factors each having 2 levels, ``TripleC'' the number of triples in the
data-set, 1 or 2 MTriples, ``Machine'', which is the software and hardware platform,
where one machine runs GNU/Linux Debian Squeeze, has 16 GB RAM, an
Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
configuration. The other machine runs Debian Wheezy, has 8 GB RAM, an
Intel Core i7-3520M CPU and a single SSD on SATA-III.

@@

To proceed beyond a visual judgment of significance, we may use
Lenth's method\todo{ref Lenth}, which is a simple method for
estimating significance in an experiment with no replicates. We plot
the results in figure~\ref{fig:fulllenth}. In
figure~\ref{fig:fullnormal}, Lenth's method is used to label the
significant effects. Visually, this is not as useful, since the large
number\todo{really include?} We see that we have very significant
effects. We also note that ``Implement'' itself is significantly
negative. We also note that the ``Implement:Lang'' interaction is
significantly positive, as expected from our setup, where ``Lang'' was
the detrimental side-effect we simulated.

To investigate whether the detrimental side-effects cancels the
positive effect of our simulated optimization, we have to formulate a
hypothesis test. By inspecting the normal plot in
figure~\ref{fig:fullnormal}, we see that only five factors are highly
significant by either being a significant main effect or participate
in a highly significant interaction. We say that the three factors
``BGPComp'', ``Range'' and ``Machine'' are \emph{inactive} since they
do not contribute significantly to the observed variation in
performance. In practical situations, the four factors ``Lang'',
``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
i.e. we cannot change the size of the data-set without loss of
information, or change the query, since it gives a different
answer. In practice, we can only control ``Implement'', i.e. we can
only change the implementation. In this context, we call ``Implement''
a \emph{control} factor and the other four active factors
\emph{environmental}.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are critical.\todo{add
  table of effects} That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. This should cause some alarm. It is a fundamental
property of two-level experiments that it can only model linear
effects. If the effect is truly linear, and the variance is small, the
two levels can be quite close together and still yield a reasonable
estimate for the effect. However, if the effect is non-linear, the
choice of levels becomes critical. In this case, the experimenter
should be concerned that they may have a non-linear effect that is not
well represented by the sizes of data-sets. Such reasoning should be
applied to all significant effects.

@@

As mentioned, the full factorial experiment goes as $2^n$ and becomes
prohibitively expensive when $n$ is a large number of
factors. As the evaluation of SPARQL endpoints is an inherently complex
problem, a large number of factors are needed, and so full experiments
cannot scale. However, we may reduce the size of the experiment
significantly by sacrificing some explanatory power, in the form of
fractional factorial experiments.

@@

For any other purpose than a rough idea of some key influences on the
overall performance of the endpoint for the given levels, this little
experiment is insufficient.

We turn to the 64-run experiment and its normal plot in
figure~\ref{fig:frac64normalalpha005}. We see that the we get more significant
effects, and that they correspond well to those we saw in the full
factorial experiment. They also all have an intuitive
explanation as above. However, we do not see the effects that we found
worrisome in the full factorial design. The ``Implement:TripleC''
interaction emerges only for a level $\alpha=0.15$, see
figure~\ref{fig:frac64normalalpha015}. To ensure that we discover such
cases will be a key challenge in further studies.

@@

The importance of finding flaws in the levels is highlighted by the
normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
experiment made it feasible to run with a larger data-set, this
experiment uses the original 15~MTriples data-set of
\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
previous discussion, for larger ``TripleC'', our simulated performance
is actually worse, as evident from the positive
``Implement''. We also see that the interactions ``Implement:Lang''
and ``Implement:TripleC'' is highly significant. It is clear what
happened: For larger ``TripleC'', the delay we introduced in the
langMatches function has a devastating effect, since it delays the
execution for every returned triple, and thereby completely dominates
the execution time.

Moreover, we note that the ``Machine'' factor has become highly
significant in the other end along with its ``Machine:TripleC''
interaction. One should take careful notice that the ``Machine''
factor encompasses many different things, that should be different
factors; CPU, disks, available RAM, installed software, tuning
parameters and so on. It is advantageous to have such broad factors in
the design in early experiments, because they may point out unexpected
effects that should be further scrutinized. The ``Machine'' factor is
a good example of usage that also helps us manage
complexity: If possible, one may group many factors into one if one is
in doubt of its relevance and break it up in subsequent
experiments. In this case, the cause of the performance problem was
highly audible: It was the old P-ATA disks of the larger machine, but
in general, such clues are not available to the experimenter.

We are forced to conclude that the three preceding experiments are
severely flawed, as they failed to take into account the non-linear
effects arising from larger data sizes on slow hardware.

Note that we do not take advantage of the usual performance of 4store
in this simulation, to the contrary, we chose 4store because of its
relative simplicity to modify it to suite our needs for illustrative
purposes, which consists of degrading it in many ways. The small
data-sets used here does not imply that the approach will not work for
large data-sets; properly configured, actual optimization will give
very different runtimes, and the small number of runs will make it
possible to evaluate a very large number of factors.

@@

\section*{Acknowledgments}
